Tests of the efficiency of various fMRI designs
===============================================

Time-stamp: <2013-01-08 17:06 christophe@pallier.org>

Note: This document is generated from the source file "simul5.Rmd", a Rmarkdown document that can be edited, e.g., with rstudio.

We present some simulations comparing the power of designs with fixed ISI to designs with varying ISI. We are interested in how precise are the estimates of the parameters associated to each condition.


```{r initialisation, echo=FALSE, results=FALSE}
rm(list=ls())
require(MASS) # for 'ginv'

# misc. useful functions

ihrf <- scan('hrf.dat') # hrf sampled at 10Hz

dirac.comb <- function(onsets, durations, total.duration, sampling.rate=10)
  # given vectors 'onsets' and 'durations' in secs, generate an indicator variable
  # sampled at 10Hz
{
  stopifnot((length(onsets)==length(durations)) | (length(durations)==1))
  if (length(durations)==1) 
    durations <- rep(durations, length(onsets))
  onsets <- ifelse(onsets<0, 0, onsets) # does not handle negative onsets
  size <- (total.duration) * sampling.rate
  u <- numeric(size)
  for (i in 1:length(onsets)) 
  {
    i1 <- round(sampling.rate*onsets[i])
    i2 <- round(sampling.rate*(onsets[i]+durations[i]))
    i1 <- min(i1, size)
    i2 <- min(i2, size)
    u[i1:i2] <- 1.0
  }
  u[1:size]
}

plot_paradigm <- function(timing)
  {
       plot(timing$onsets, 
            rep(1,nrow(timing)), 
            col=timing$conditions,
            type='h', 
            ylim=c(0,1.5), 
            ylab='', 
            xlab='Time',
            main="conditions")
  }

hrf <- function(onsets, durations, total.duration, TR=1)
  # convolution of a time series (defined by onsets & durations of 'boxcar' events) by impulse hrf
{
  stopifnot(TR==1) # undersampling not yet implemented
  if (!exists("ihrf")) 
    ihrf <- scan('hrf.dat') # hrf sampled at 10Hz
  x2 <- c(rep(0, length(ihrf)), dirac.comb(onsets, durations, total.duration))
  x3 <- stats::filter(x2, ihrf, sides=1)
  x4 <- x3[-(1:length(ihrf))]
  x4 # here we should do some undersampling if TR>1
}

create_design_matrix <- function(timing, totalduration)
# timing must be a data.frame with columns 'conditions', 'onsets', and 'durations' 
  {
  stopifnot(is.data.frame(timing))
  condnames <- sort(unique(timing$conditions))
  ncond <- length(condnames)
  regressors <- list()
  for (cond in 1:ncond)
    {
    select <- timing$conditions==condnames[cond]
    regressors[[cond]] <- scale(hrf(timing$onsets[select],
                                   timing$durations[select],
                                   totalduration), 
                               center=TRUE,
                               scale=FALSE)
    }
  names(regressors) <- condnames
  as.matrix(as.data.frame(regressors))
  }

plot_design_matrix <- function(X)
  {
  matplot(1:nrow(X), X, 
          type='l', lty=1, col=1:ncond)
  }

```


``` {r efficiency, echo=FALSE}
efficiency <- function(contrast, designMat)
  # returns the efficiency of a contrast for a given design 
  # according to Henson (2007), formula 15.4 (from Dale, 1999)
  {
  sum(diag(ginv(t(contrast) %*% ginv(crossprod(designMat)) %*% contrast)))
  }


```

We define various ways of generating paradigms: 

```{r paradigms}
generate_paradigm_fixed_SOA <- function(ncond, trialpercond,  stimduration, SOA, totalduration)
  # ncond os the number of different experimental conditions
  # stimduration, SOA and totalduration must be expressed in seconds
  {
    ntrials <- ncond * trialpercond
    conditions <- sample(rep(1:ncond,trialpercond))
    onsets <- (1:ntrials-1)*SOA
    durations <- rep(stimduration, ntrials)
    timing <- data.frame(onsets, conditions, durations)
    timing
  }

generate_paradigm_fixed_SOA_adding_silences <- function(ncond, trialpercond,  stimduration, SOA, totalduration0)
  {
  # we simply add one more condition 'silence' having as many trials as the others,
  # lengthening the whole duration by 'trialpercond*SOA'
    timing <- generate_paradigm_fixed_SOA(ncond+1, trialpercond, stimduration, SOA, totalduration0+ trialpercond*SOA)
  # and delete the references to it (it will be in the implicit baseline)  
    timing <- subset(timing, conditions != ncond+1)
    timing
    }

generate_paradigm_varying_SOA <- function(ncond, trialpercond,  stimduration, SOA, jitter, totalduration)
  {
    # first, we computed a fixed SOA design 
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
    ntrials <- nrow(timing)
    # introducing some random jittering around '0' between the trials
    timing$onsets <- abs(timing$onsets+runif(ntrials, min=-jitter/2, max=jitter/2))
    timing
  }

generate_paradigm_varying_SOA_null_events  <- function(ncond, trialpercond,  stimduration, SOA, totalduration)
  {
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
# insert silences. This will lengthen the whole by trialpercond*SOA
    totalsilence <- trialpercond*SOA
    totalduration <- totalduration+totalsilence
    ntrials <- ncond*trialpercond   
    silences <- runif(ntrials, 0.2,1.8) * (totalsilence / ntrials)
    silences <- silences * (totalsilence/sum(silences))
    timing$onsets <- timing$onsets+cumsum(silences)
    timing
  }
```



```{r fig.width=7, fig.height=6}
ncond <- 5
trialpercond <- 20
stimduration <- 6
SOA <- 10
(totalduration <- (ncond * trialpercond * SOA) + SOA)
````

We can generate a paradigm with a fixed SOA...

```{r}
timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)

head(timing)
```

... and build its associated design matrix

```{r generateDesignMat, fig.width=7, fig.height=6}
X <- create_design_matrix(timing, totalduration)
par(mfrow=c(2,1))
plot_paradigm(timing)
plot_design_matrix(X)
```


Simulations and Estimation by an hrf model. 
------------------------------------------

For many such designs, we simulate a voxel where the signal increases in a linear fashion with 'condition' (amplitudes=1:5).


```{r simul_fun}

normalnoise <- function(npoints) { 5*rnorm(npoints) }

simulations <- function(nsim, beta, listofcontrasts, noise_fun, design_generator, ...)
  {
  estimates <- matrix(nrow=nsim, ncol=length(beta))
  estimatese <- matrix(nrow=nsim, ncol=length(beta))
  efficiencies <- matrix(nrow=nsim, ncol=length(listofcontrasts))
  
  for (i in 1:nsim)
    {
      timing <- design_generator(...)
      X <- create_design_matrix(timing, totalduration)
      
      # simulate a time-series 
      y0 <- X %*% beta
      y <- y0 + noise_fun(length(y0))
      
      # analyse using a linear model
      m <- summary(lm(y~ X))
      estimates[i,] <- m$coefficients[-1,1] # get the estimations of the parameters
      estimatese[i,] <- m$coefficients[-1,2] # and their associated standard errors

      # compute the efficiency of contrasts 
      for (contrast in 1:length(listofcontrasts)) {
          efficiencies[i,contrast] <- efficiency(listofcontrasts[[contrast]],X)
          }

      }
  # returns the estimates, their s.e. and the efficiencies
  list(estimates=estimates, 
       estimatese=estimatese, 
       efficiencies=efficiencies)
}
  
```

```{r simul_fixed_SOA}
nsim <- 10 # number of simulations

betas <- 1:ncond # theoretical amplitudes (per condition)

con <- betas-mean(betas) # contrast of interest
con %*% betas # value of the contrast to be estimated

estimates <- matrix(nrow=nsim, ncol=ncond) 
estimatese <- matrix(nrow=nsim, ncol=ncond) 

conestimates <- numeric(nsim)
eff <- numeric(nsim)
beta1eff <- numeric(nsim)

for (sim in 1:nsim)
{
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
    X <- create_design_matrix(timing, totalduration)
    eff[sim] <- efficiency(con, X)
    beta1eff[sim] <- efficiency(c(1,rep(0,length(betas)-1)),X)    
    y0 <- X %*% betas 
    noise <- 5*scale(rnorm(length(y0)))
    y <- y0 + noise
    summarylmmod <- summary(lm(y~X))
    estimates[sim,] <- summarylmmod$coefficients[-1,1]
    estimatese[sim,] <- summarylmmod$coefficients[-1,2]
    conestimates[sim] <- con %*% estimates[sim,]
}
```

Estimation of the last model of the simulation
```{r lastmod}
summary(lm(y~X))
```

Summaries of estimates of betas and their standard errors
```{r reportestimates}
summary(estimates)
summary(estimatese)
```


```{r plots, fig.width=7, fig.height=6}
par(mfrow=c(2,2))
# distributions of estimates of individual betas
boxplot(estimates, main='estimates of the betas')
grid()
boxplot(beta1eff, main='efficiency for beta1')
grid()

# distribution of estimate of the contrasts
boxplot(conestimates, main='estimates of the contrast')
grid()
boxplot(eff, main='efficiency for linear contrast')
grid()

```

The average standard deviation of the estimates of the betas is:

```{r}
mean(apply(estimates,2,sd))
```

Remark that the variability in the estimates depends: (a) on the variability of the design matrices and (b) on the noise. 


```{r simulfun_fixed_SOA}
o <- simulations(
  nsim=10,
  beta=1:ncond,
  listofcontrasts=list((1:5)-2.5,c(1,0,0,0,0)),
  normalnoise,
  generate_paradigm_fixed_SOA, 
  ncond, trialpercond, stimduration, SOA, totalduration)

par(mfcol=c(1,2))
boxplot(o$estimates)
mean(unlist(o$estimatese))
boxplot(o$efficiencies)
```

New design, with jitter between trials:
--------------------------------------

Now, we jitter the SOA between trials and run a similar simulation.

```{r}
timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
ntrials <- nrow(timing)
jitter <- 4
timing$onsets <- abs(timing$onsets+runif(ntrials,min=-jitter/2, max=jitter/2))
head(timing)
par(mfcol=c(1,1))
hist(diff(timing$onsets, main='Distribution of SOAs'))
```

```{r generateDesignMat}
```

```{r simul_varying_SOA, fig.width=7, fig.height=6}
#nsim <- 100 # number of simulations
betas <- 1:ncond # theoretical amplitudes (per condition)
con <- betas - mean(betas)
con %*% betas # value of the contrast to be estimated

estimates <- matrix(nrow=nsim, ncol=ncond) 
estimatese <- matrix(nrow=nsim, ncol=ncond) 

eff <- numeric(nsim)
for (sim in 1:nsim)
{
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
    ntrials <- nrow(timing)
    # introducing some random jittering
    timing$onsets <- abs(timing$onsets+runif(ntrials, min=-jitter/2, max=jitter/2))
    X <- create_design_matrix(timing, totalduration)
    eff[sim] <- efficiency(con, X)
    beta1eff[sim] <- efficiency(c(1,rep(0,length(betas)-1)),X)  
    y0 <- X %*% betas 
    noise <- 5*scale(rnorm(npoints))
    y <- y0 + noise
    summarylmmod <- summary(lm(y~X))
    estimates[sim,] <- summarylmmod$coefficients[-1,1]
    estimatese[sim,] <- summarylmmod$coefficients[-1,2]
    
    conestimates[sim] <- con %*% estimates[sim,]
}
```

```{r lastmod}
```

Summaries of estimates of betas and their standard errors
```{r reportestimates}
```


```{r, fig.width=7, fig.height=6}
par(mfrow=c(2,2))
# distributions of estimates of individual betas
boxplot(estimates, main='estimates of the betas')
grid()
boxplot(beta1eff, main='efficiency for beta1')
grid()

# distribution of estimate of the contrasts
boxplot(conestimates, main='estimates of the contrast')
grid()
boxplot(eff, main='efficiency for linear contrast')
grid()

```

The average standard deviation of the estimates of the betas is:

```{r}
mean(apply(estimates,2,sd))
```

```{r simulfun_varying_SOA}

timing <- generate_paradigm_varying_SOA(ncond, trialpercond, stimduration, 4, SOA, totalduration)
plot_paradigm(timing)
X <- create_design_matrix(timing, totalduration)
plot_design_matrix(X)

o <- simulations(
  nsim=10,
  beta=1:ncond,
  listofcontrasts=list((1:5)-2.5,c(1,0,0,0,0)),
  normalnoise,
  generate_paradigm_varying_SOA, 
  ncond, trialpercond, stimduration, 4, SOA, totalduration)

par(mfcol=c(1,2))
boxplot(o$estimates)
mean(unlist(o$estimatese))
boxplot(o$efficiencies)
```


First conclusion
----------------

The efficiency for the linear contrast is about the same for the two types of design, but individual betas are better estimated with jittering. But we have not included null events yet.  

Simulations with 'explicit' NULL events
---------------------------------------

Coming back to the fixed SOA schedule, we now add a sixth condition, consisting of silent trials of the same duration as the 'real' trials.

```{r simul_fixed_SOA_silence}
ncond <- 6
trialpercond <- 20
stimduration <- 6
SOA <- 10
(totalduration <- (ncond * trialpercond * SOA) + SOA)


#nsim <- 100 # number of simulations
betas <- c(1:5,0)  # theoretical amplitudes (per condition)

con <- c((1:5)-mean(1:5),0)  # contrast of interest
con %*% betas # value of the contrast to be estimated

estimates <- matrix(nrow=nsim, ncol=ncond) 
estimatese <- matrix(nrow=nsim, ncol=ncond) 
conestimates <- numeric(nsim)
eff <- numeric(nsim)
beta1eff <- numeric(nsim)

for (sim in 1:nsim)
{
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration)
    X <- create_design_matrix(timing, totalduration)
    eff[sim] <- efficiency(con, X)
    beta1eff[sim] <- efficiency(c(1,rep(0,length(betas)-1)),X)    
    y0 <- X %*% betas 
    noise <- 5*scale(rnorm(length(y0)))
    y <- y0 + noise
    summarylmmod <- summary(lm(y~X))
    estimates[sim,] <- summarylmmod$coefficients[-1,1]
    estimatese[sim,] <- summarylmmod$coefficients[-1,2]
    conestimates[sim] <- con %*% estimates[sim,]
}
```

```{r lastmod}
```

Summaries of estimates of betas and their standard errors
```{r reportestimates}
```


```{r, fig.width=7, fig.height=6}
par(mfrow=c(2,2))
# distributions of estimates of individual betas
boxplot(estimates, main='estimates of the betas')
grid()
boxplot(beta1eff, main='efficiency for beta1')
grid()

# distribution of estimate of the contrasts
boxplot(conestimates, main='estimates of the contrast')
grid()
boxplot(eff, main='efficiency for linear contrast')
grid()

```

The average standard deviation of the estimates of the betas is:

```{r}
mean(apply(estimates,2,sd))
```


And now, the same total amount of silence is inserted as NULL events of varying length between the 'real' trials, introducing jitter.

```{r simul_varying_SOA_with_NULL_events}
ncond <- 5
trialpercond <- 20
stimduration <- 6
SOA <- 10
(totalduration0 <- (ncond * trialpercond * SOA) + SOA)

betas <- 1:ncond
con <- betas-mean(betas) # contrast of interest
con %*% betas # value of the contrast to be estimated

#nsim <- 100
estimates <- matrix(nrow=nsim, ncol=ncond) 
estimatese <- matrix(nrow=nsim, ncol=ncond) 
conestimates <- numeric(nsim)
eff <- numeric(nsim)
beta1eff <- numeric(nsim)

for (sim in 1:nsim)
{
    timing <- generate_paradigm_fixed_SOA(ncond, trialpercond, stimduration, SOA, totalduration0)
# insert silences
    totalsilence <- trialpercond*SOA
    totalduration <- totalduration0+totalsilence
    ntrials <- ncond*trialpercond   
    silences <- runif(ntrials, 0.2,1.8) * (totalsilence / ntrials)
    silences <- silences * (totalsilence/sum(silences))
    timing$onsets <- timing$onsets+cumsum(silences)
      
    X <- create_design_matrix(timing, totalduration)
    eff[sim] <- efficiency(con, X)
    beta1eff[sim] <- efficiency(c(1,rep(0,length(betas)-1)),X)    
    y0 <- X %*% betas 
    noise <- 5*scale(rnorm(length(y0)))
    y <- y0 + noise
    summarylmmod <- summary(lm(y~X))
    estimates[sim,] <- summarylmmod$coefficients[-1,1]
    estimatese[sim,] <- summarylmmod$coefficients[-1,2]
    conestimates[sim] <- con %*% estimates[sim,]
}
```

```{r lastmod}
```

Summaries of estimates of betas and their standard errors
```{r reportestimates}
```


```{r, fig.width=7, fig.height=6}
par(mfrow=c(2,2))
# distributions of estimates of individual betas
boxplot(estimates, main='estimates of the betas')
grid()
boxplot(beta1eff, main='efficiency for beta1')
grid()
mean(apply(estimates,2,sd))

# distribution of estimate of the contrasts
boxplot(conestimates, main='estimates of the contrast')
grid()
boxplot(eff, main='efficiency for linear contrast')
grid()

```

Conclusions
-----------

For the linear contrast, the two two types of designs provide similar power. For the estimation of the betas, the paradigms with varying ISI perform better.

However, our simulations looked at the general behavior of the paradigm types. We can select the best schedules from a class of paradigm.

